---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->

[![Codecov test coverage](https://codecov.io/gh/Polymerase3/phiperio/graph/badge.svg)](https://app.codecov.io/gh/Polymerase3/phiperio)
[![R-CMD-check](https://github.com/Polymerase3/phiperio/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Polymerase3/phiperio/actions/workflows/R-CMD-check.yaml)
[![pkgcheck](https://github.com/Polymerase3/phiperio/workflows/pkgcheck/badge.svg)](https://github.com/Polymerase3/phiperio/actions?query=workflow%3Apkgcheck)
[![test-coverage](https://github.com/Polymerase3/phiperio/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/Polymerase3/phiperio/actions/workflows/test-coverage.yaml)
[![version](https://img.shields.io/github/r-package/v/Polymerase3/phiperio?label=version)](https://github.com/Polymerase3/phiperio)
[![Project Status: Active/Unstable](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- badges: end -->

# phiperio

The `phiperio` package provides utilities to import, validate, and manage PhIP-Seq datasets, including standardized conversion pipelines, data checks, and access to cached peptide metadata.

## Installation

You can install the development version of `phiperio` from GitHub with either `pak` or `devtools`:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("Polymerase3/phiperio")

# or, using devtools:
# install.packages("devtools")
devtools::install_github("Polymerase3/phiperio")
```

## Usage

For guided walk-throughs, see the pkgdown vignettes:

- **[Importing long tidy data (convert_standard)](https://polymerase3.github.io/phiperio/articles/importing-long-tidy-data.html)** — cross-sectional and longitudinal tidy inputs.
- **[Importing multiple files at once](https://polymerase3.github.io/phiperio/articles/importing-multiple-files.html)** — batch ingest of many CSV/Parquet files with `sample_id_from_filenames`.
- **[Importing legacy PhIP-Seq data (convert_legacy)](https://polymerase3.github.io/phiperio/articles/importing-legacy-data.html)** — classic wide matrices (exist/fold_change/raw counts) plus sample/timepoint metadata.

## Aim and key features

`phiperio` focuses on reliable ingest and validation of PhIP-Seq data, so downstream analyses start from a clean, standardized base. Key features include:

- **DuckDB backend + Parquet first:** uses DuckDB under the hood and writes/reads Parquet by default as the transaction layer between the `phiper` data source and `phiperio`, giving fast I/O and great interoperability.
- **Scales to millions of rows:** lazy database pipelines and Parquet storage let you work efficiently with very large PhIP-Seq datasets.
- **Import helpers** for common PhIP-Seq inputs and peptide metadata (peptide library cached and maintained in the companion `phiper` repo).
- **Strong validation** and consistency checks to catch data issues early.
- **Lightweight, reproducible pipelines** to standardize raw inputs into `<phip_data>` objects.

## Issues

Spotted a bug or want to request a feature? Please open an issue: https://github.com/Polymerase3/phiperio/issues
